---
title: "Using Stratified Cox Regression to Test Competing Risks"
author: "Boyi Guo"
institute: |
  | Department of Biostatistics
  | University of Alabama at Birmingham
date: "Nov 19, 2020"
output: 
  beamer_presentation:
    theme: "Szeged"
    colortheme: "spruce"
    toc: FALSE
    number_section: false
    slide_level: 2
bibliography:
  - bibliography.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
library(drake)
library(survival)
library(gtsummary)
library(tidyverse)
loadd(analytic_dat)
loadd(joint_cs_mdl)
loadd(strk_cs_mdl)
loadd(chd_cs_mdl)
```


# Objectives
- Introduce an analytic approach to jointly model the cause-specific hazard for two or more competing events 
  - Specifically, the hypothesis testing of differential effects <!-- of an exposure --> on the competing events 
  <!-- where I will review what competing event is later -->
- Demonstrate the implementation of the Cause-specific Joint Model

<!-- A quick disclaimer here: I am focusing on the application perspective of the method, rather than the theoretical perspective. I will defer the theoretical discussion to other experts in the group. -->

# Outline
- Survival Analysis Review
  - Stratified Cox Regression
  - Competing Risk Analysis
  - Cause-specific Hazard
- Cause-specific Hazard Joint Modeling
  - Problem Statement
  - Formulation
  - Implementation
- REGARDS Example
- Conclusion


# Survival Analysis Review

Definition:
Survival analysis is a way to model time-to-event outcome while accounting for (mostly right) censoring that is non-informative and random.

  - censoring: events of interest are not observed over the course of a study

<!-- Example: people who died of COVID(unfornately) would considered as censored when we use REGARS data to study cardiovascular outcomes. or By the time REGARDS follow-up phases ends, people who still experience cardiovascular events would be consdidered censored-->

Components:

- Outcome: time-to-event and censoring status $(T_i, C_i)$
- Survival function: $S(t) = Pr(T > t)$
<!-- Survival function descirbes the distribution of event times. Kaplan-meier plot: the probability of the event of interest happens later than a time t.-->
- Cumulative incidence function: $F(t) = 1- S(t) = Pr(T \le t)$
- Hazard function:
$$
h(t) = \lim\limits_{\Delta t \rightarrow0}\frac{Pr(t\leq T < t+\Delta t|T \geq t)}{\Delta t}
$$

<!-- Hazard describes  the instantaneous rate of the occurrence of the event of interest in subjects who are currently at risk of the event-->
<!-- with enough arithmetic manipulation, you can find a one-to-one relationship between h(t) and S(t), and hence can infer the survival function based on hazard function-->

## Cox Propotional Hazard Regression
- Assumption: the hazard are proportional
    <!-- Fast and dirty way to evaluate is by looking at Kaplan Meier. No cross over-->
- Formulation:
  $$
    h(t|x_i) = h_0(t)\exp(\beta x_i)
  $$

  where $\exp(\beta)$ is the hazard ratio,    
  $h_0(t)$ is the baseline hazard function
- Hazard ratio (HR) ranges (0,$\infty$), 
  - HR > 1 means more likely to observe the event <!--compared to the reference value/level-->
  - HR < 1 means less likely to observe the event
    

## Stratified Cox Proportional Hazard Regression   
  - Used when stratified analysis are required
  - Additional assumption: the risk associated with an exposure are the same across all strata, 
  - Formulation:
    $$
      h_j(t|x_i) = h_{0j}(t) \exp(\beta x_i),
    $$
    where $h_{0j}(t)$ is the baseline hazard function for the $j$ the strata.
    $exp(\beta)$ is the hazard ratio across all strata
    
## Stratified Cox Regression Example

Race stratified analysis for CVD outcomes

  - Research question: what is the effect of smoking on CVD risk among Blacks and Whites
  - The hazard functions for different race groups are different because of the distinct baseline hazard function
  - The estimate of hazard ratio associated with smoking is exp($\beta$) for all race groups

## Competing Risk Analysis
Definition: Competing risk exists when we have two or more mutually exclusive events of consideration

  - either stroke or coronary heart disease (CHD)
<!-- The reason of why we need a specific statistical models for competing risk is because observing one of the events change the risk of the other events-->
<!-- The reason being observing one of the events change the risk of the other events-->
<!-- TODO: improve this definition -->

Components:

  - Outcome:($T_i, C_i, J_i)$, where $J_i\in \{1, 2(, \dots) \}$ for event types
  - Two types of hazard
    - Cause-specific hazard
    - Sub-distribution hazard, AKA Fine-Grey method

@austin2016introduction stated "_subdistribution hazards models should be used for developing clinical prediction models and risk-scoring systems for survival outcomes, whereas cause-specific hazard models may be more appropriate for addressing epidemiological questions of etiology._"


## Cause-specific hazard models
- Used when studying etiology of diseases [@lau2009competing]
- Fitting Cox models by focusing on the events of primary interest and treating the competing events as censored.
- Formulation:
  $$h_{J}(t|x_i)=h_{0,J}(t)\exp(\beta_{J} x_i)$$
  where $J$ denotes different types of events

- Animated demonstration in PowerPoint

# Cause-specific Hazard Joint Modeling

## REGARDS Example
- Events of Interest: Stroke / CHD
- Exposure of Interest: Smoking status (at baseline)
- Research Question: 
  - How does smoking affect the risk of stroke considering competing risk of CHD?
  $$h_{Stroke}(t|x_i)=h_{0,Stroke}\exp(\beta_{Stroke} x_i)$$
  - How does smoking affect the risk of CHD?
  $$h_{CHD}(t|x_i)=h_{0,CHD}\exp(\beta_{CHD} x_i)$$
  - Does smoking affect the risk of stroke the same way as the risk of CHD


## Problem Statement
How to test if the impact on the risks from an exposure are the same for all events?

- The original cause-specific hazard model requires two models, one for each type of events by treating the other as censored outcome
- Statistically, the hazard ratio of exposures are not directly comparable


## Cause-specific Hazard Joint Model

- @lunn1995applying proposed a joint modeling approach to estimate the cause specific hazards for both events using stratified Cox regression
<!-- The idea is really similar to using statistical interaction to address modifying effect, which firstly analyzed using stratified analysis-->
- Formulation for two event types:
<!-- $$ -->
<!-- h_{CHD}(t|x_i)=h_{0,CHD}\exp(\beta x_i + \beta^\prime\delta_i x_i), -->
<!-- $$ -->
<!-- $$ -->
<!-- h_{Stroke}(t|x_i)=h_{0,Stroke}\exp(\beta x_i + \beta^\prime\delta_i x_i), -->
<!-- $$ -->
$$h_{J}(t|x_i)=h_{0J}(t)\exp(\beta x_i + \beta^\prime\delta_i x_i)$$
where $\delta$ is an indicator variable: $\delta_i=1$ if J=1 and $\delta_i=0$ if J=2

<!-- Unlike the original Cause-specific Model, we have one model that could model both types, and hence the risk "difference" in multiplicative scale is statistically testable. -->

## Implementation
- Basic idea:
  1. Data augmentation
      - Duplicate each event (both types) by treating the duplication as censored outcome
      - Duplicate the censored events
      - Create interaction term of event type & exposure
  2. Stratified Cox model based on event type
    $$h_{CHD}(t|x_i)=h_{0,CHD}\exp(\beta x_i + \beta^\prime\delta_i x_i)$$
    $$h_{Stroke}(t|x_i)=h_{0,Stroke}\exp(\beta x_i + \beta^\prime\delta_i x_i)$$
  
  3. Test the interaction term $\beta^\prime$
  
  
- Animated demonstration in PowerPoint

<!-- ### Modified Cox PH Models - Lunn & McNeil -->
<!-- - Data Augmentation -->
<!-- - Stratified Model Based on event type -->


# Toy Example
- Events of Interest: Stroke / CHD
- Exposure of Interest: Smoking status (at baseline)
- Covariates: Age, Race
- Research Question: 
  - How does smoking affect the risk of stroke considering competing risk of CHD?
  - How does smoking affect the risk of CHD?
  - Does smoking affect the risk of stroke the same way as the risk of CHD


## Cause-specific Hazard for Stroke
```{r}
tbl_regression(strk_cs_mdl, exponentiate = TRUE,
               label = list(Age ~ "Age (years)", Race ~ "Race (Whites Vs Blacks)", Smoke_current ~ "Current Smoker"),
               show_single_row = Race) %>% 
  italicize_levels() #%>% 
  # bold_labels()
```

## Cause-specific Hazard for CHD
```{r}
tbl_regression(chd_cs_mdl, exponentiate = TRUE,
               label = list(Age ~ "Age (years)", Race ~ "Race (Whites Vs Blacks)", Smoke_current ~ "Current Smoker"
                            ),
                show_single_row = c(Race)) %>% 
  italicize_levels() #%>% 
  # bold_labels()
```

## Joint Model
```{r}
tbl_regression(joint_cs_mdl, exponentiate = TRUE,
               label = list(Age ~ "Age", Race ~ "Race (Whites Vs Blacks)", Smoke_current ~ "Current Smoker",
                            intr_Race ~ "Stroke*Race( Whites Vs Blacks)",
                            intr_smoke ~ "Stroke*Smoker",
                            intr_age ~ "Stroke*Age"),
                show_single_row = c(Race,intr_Race)) %>% 
  italicize_levels() #%>% 
  # bold_labels()
```

## Experience
- Include the interaction for all variables in the model, including strata-covariates interactions
- Need extra cautious when interpreting the cause-specific hazard ratios
  - Understand which strata is the reference level
- Possible to extend to exposures that are more than 2 levels, or more than two competing events
  - Extra effort to setup the design matrix, i.e. the interaction terms


# Conclusion
- Review the basic concepts about survival analysis and competing risk analysis
  - Cause-specific hazard, preferably used in epidemiological study
- Cause-specific hazard joint model proposed by @lunn1995applying
  - Simultaneously models two or more competing events
  - Stratified Cox Regression with a data augmentation step
  - Test the effect change in competing risks via statistical interaction
- Demonstrate the implementation
  - Results are the same <!-- for the marginal CS hazard and Joint CS hazard, with additional hypothesis testing for effect change-->
  - Cautious about the reference level <!-- in interpretation-->

## Ackowledgement
- Dr. George Howard <!-- finding such interest paper & offers the oppurtunity to let me lead this discussion-->
- Dr. D. Leann Long <!-- valuable feedbacks on improving the presentation-->
- Vicky Wang <!-- my girlfriend who fed me in the past couple days to make me still alive today-->
- Rachel Stuckwisch <!-- tolerate me ingoring her texts-->

## Reference


